### 2020-04-01 06b-bisimulation.tex
Update 06b-bisimulation.tex
examples + flat fucntion
fino a sect4 e flat
update fino a sezione 4 inclusa e def di flat
just typos

For this reason, we propose an alternative notion of bisimulation, that hereafter we call \emph{biosimulation},
that allows us  to compare two biological systems by restricting the observation to only the limited events of interest.
This allows one to tailor the equivalence to different applications and purposes.

@@

To simplify the query language, we first introduce a function that takes a solid link chain and returns a simple string by eliminating all the channel matching pairs leaving just one channel per each pair. This transformation is harmless, in the sense that it retains all the information in the chain, because it is applied to solid chains only. The function $\mathit{flat}$ is defined inductively as follows:
\[
\begin{array}{lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl}
\mathit{flat}(\epsilon)&\defeq & \epsilon &
\mathit{flat}(\link{\silent}{\beta}) & \defeq & \beta &
\mathit{flat}(\link{\alpha}{\silent})& \defeq & \epsilon &
\mathit{flat}(\link{\alpha}{\beta}) & \defeq &\beta \\
\end{array}\]
\[
\begin{array}{lcl}
\mathit{flat}(\link{\alpha}{\beta}\upsilon)& \defeq & \beta :: \mathit{flat}(\upsilon)\\
\end{array}\\
\]
where  $::$ is the usual string concatenation operator.

@@

\begin{definition}[Semantics of assertions]
 We define $\llbracket F\rrbracket \subseteq\wp(\mathit{names}^*)$ by induction on the structure of $F$: 
\[
\begin{array}{rcl}
 \llbracket \alpha \rrbracket &\defeq&  \{\alpha\} \\
 \llbracket ? \rrbracket &\defeq&   \mathit{names}\\
 \llbracket [N]\rrbracket &\defeq&   N\\
 \llbracket \epsilon \rrbracket &\defeq&  \{\epsilon\} \\
 \llbracket F_1::F_2 \rrbracket &\defeq& \{\omega_1 :: \omega_2~\mid~ \omega_1 \in \llbracket F_1 \rrbracket \wedge \omega_2 \in \llbracket F_2 \rrbracket\}  \\
 \llbracket F^+ \rrbracket &\defeq& \llbracket F \rrbracket^+  \\
 \llbracket F^* \rrbracket &\defeq& \llbracket F \rrbracket^*  \\
 \llbracket  F_1 \vee F_2 \rrbracket &\defeq& \llbracket  F_1\rrbracket \cup \llbracket  F_2\rrbracket  \\
 \llbracket  F_1 \wedge F_2 \rrbracket &\defeq& \llbracket F_1 \rrbracket \wedge \llbracket F_2\rrbracket 
\end{array}
\]
\end{definition}

@@

The notion of biosimulation builds on the above language of assertions to parameterize the induced equivalence on the property of interest. 
Please recall that we have defined the behaviour of the contest in a non determinist way, thus 
at each step, different possible sets of entities can be provided to the system and different sets of reaction can be enabled/disabled. 
Biosimulation can thus be used to compare the behaviour of different systems that share some of the reactions or entities or also to compare the behaviour of the same set of reaction rules when different contexts are provided.

@@

\begin{definition}[Biosimilarity  $ \sim_F $]
Given an assertion $F$, a \emph{biosimulation} $\mathbf{R}_F$ that respects $F$ is a binary relation over {\tt link}-calculus processes such that, if $P \mathrel{\mathbf{R}_F} Q$ then:
 \begin{itemize} 
 \item
 for any $v,P'$ such that 
 $P \xrightarrow{v} P'$ then there exist $v',Q'$ such that $Q  \xrightarrow{w} Q'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\item
 for any $w,Q'$ such that 
 $Q \xrightarrow{w} Q'$ then there exist $v,P'$ such that $P  \xrightarrow{v} P'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\end{itemize}
%\item i$Q \xrightarrow{v} Q'$ with $v \entails F$ iff $\exists v',P'$ s.t.~$v \stretcheq v'$, $P  \xrightarrow{v'} P'$ with $v \entails F$. and $P' \mathrel{\mathbf{R}_{F}} Q'$.
%\end{itemize}
We let $ \sim_F $ denote the largest  biosimulation and we say that $P$ is \emph{biosimilar} to $Q$, with respect to $F$, if $P \sim_F Q$.
\end{definition}

\begin{remark}
An alternative way to look at a biosimulation that respects $F$ is to define it as an ordinary bisimulation over the transition system labelled over $\{F,\neg F\}$ obtained by transforming each transition $P \xrightarrow{v} P'$ such that $v\entails F$ into $P \xrightarrow{F} P'$ and  each transition $P \xrightarrow{v} P'$ such that $v\not\entails F$ into $P \xrightarrow{\neg F} P'$.
\end{remark}

@@

Finally, we extend the classical result establishing the correspondence between the logical equivalence induced by HML with bisimilarity for proving that biosimilarity coincides with bio-logical equivalence.

>>>>

For this reason, we propose an alternative notion of bisimulation, that hereafter we call \emph{bio-simulation},
that allows us  to compare two biological systems by restricting the observation to only the limited events of interest.
This allows one to tailor the equivalence to different applications and purposes.

@@

To simplify the query language, we first introduce a function that takes a solid link chain and returns a simple string by eliminating all the channel matching pairs leaving just one channel per each pair. This transformation is harmless, in the sense that it retains all the information in the chain, because it is applied to solid chains only. Below we denote by $\mathit{decs} = \{ s, \overline{s}, \widetilde{s},\widehat{s},\underline{s}~|~ s\in S\}$ the set of decorated names (without subscripts $_i$ and $_o$).
The function $\mathit{flat}$ is defined inductively as follows:
%\[
%\begin{array}{lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl@{\hspace{0.5cm}}lcl}
%\mathit{flat}(\epsilon)&\defeq & \epsilon &
%\mathit{flat}(\link{\gamma}{\gamma'}) &  \defeq & \left \{ \begin{array}{cl} \epsilon & \mathrm{if } (\gamma = \beta_i \wedge \gamma = \beta_o)  \vee \gamma = \silent\\
%\gamma' & \mathrm{otw }  \end{array} \right. &
%%\mathit{flat}(\link{\alpha}{\silent})& \defeq & \epsilon &
%%\mathit{flat}(\link{\alpha}{\beta}) & \defeq &\beta \\
%\end{array}\]

\[
\mathit{flat}(\epsilon)\defeq \epsilon 
\qquad
\mathit{flat}(\link{\alpha}{\beta}) \defeq 
\left \{ \begin{array}{ll} 
\beta & \mbox{if $\beta\in \mathit{reacts}\cup\{\mathit{cxt}\}\cup\mathit{prods}$}\\
\gamma & \mbox{if $\beta=\gamma_i$ with $\gamma\in\mathit{decs}$}\\
\epsilon & \mbox{otherwise}  
\end{array} \right. 
\]

\[
\begin{array}{lcl}
\mathit{flat}(\link{\alpha}{\beta}\upsilon)& \defeq & \mathit{flat}(\link{\alpha}{\beta}):: \mathit{flat}(\upsilon)\\
\end{array}\\
\]
where  $::$ is the usual string concatenation operator.

@@

\begin{definition}[Semantics of assertions]
 We define $\llbracket F\rrbracket \subseteq\wp(\mathit{names}^*)$ by induction on the structure of $F$: 
\[
\begin{array}{rcl}
 \llbracket \alpha \rrbracket &\defeq&  \{\alpha\} \\
 \llbracket ? \rrbracket &\defeq&   \mathit{names}\\
 \llbracket [N]\rrbracket &\defeq&   N\\
 \llbracket \epsilon \rrbracket &\defeq&  \{\epsilon\} \\
 \llbracket F_1::F_2 \rrbracket &\defeq& \{\omega_1 :: \omega_2~\mid~ \omega_1 \in \llbracket F_1 \rrbracket \wedge \omega_2 \in \llbracket F_2 \rrbracket\}  \\
 \llbracket F^+ \rrbracket &\defeq& \llbracket F \rrbracket^+  \\
 \llbracket F^* \rrbracket &\defeq& \llbracket F \rrbracket^*  \\
 \llbracket  F_1 \vee F_2 \rrbracket &\defeq& \llbracket  F_1\rrbracket \cup \llbracket  F_2\rrbracket  \\
 \llbracket  F_1 \wedge F_2 \rrbracket &\defeq& \llbracket F_1 \rrbracket \cap \llbracket F_2\rrbracket 
\end{array}
\]
\end{definition}

@@

\subsection{Bio-similarity and bio-logical equivalence}

The notion of bio-simulation builds on the above language of assertions to parameterize the induced equivalence on the property of interest. 
Please recall that we have defined the behaviour of the contest in a non determinist way, thus 
at each step, different possible sets of entities can be provided to the system and different sets of reaction can be enabled/disabled. 
Bio-simulation can thus be used to compare the behaviour of different systems that share some of the reactions or entities or also to compare the behaviour of the same set of reaction rules when different contexts are provided.

@@

\begin{definition}[Bio-similarity  $ \sim_F $]
Given an assertion $F$, a \emph{bio-simulation} $\mathbf{R}_F$ that respects $F$ is a binary relation over {\tt link}-calculus processes such that, if $P \mathrel{\mathbf{R}_F} Q$ then:
 \begin{itemize} 
 \item
 for any $v,P'$ such that 
 $P \xrightarrow{v} P'$ then there exist $v',Q'$ such that $Q  \xrightarrow{w} Q'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\item
 for any $w,Q'$ such that 
 $Q \xrightarrow{w} Q'$ then there exist $v,P'$ such that $P  \xrightarrow{v} P'$ with $v\equiv_F w$ and $P' \mathrel{\mathbf{R}_F} Q'$.
\end{itemize}
%\item i$Q \xrightarrow{v} Q'$ with $v \entails F$ iff $\exists v',P'$ s.t.~$v \stretcheq v'$, $P  \xrightarrow{v'} P'$ with $v \entails F$. and $P' \mathrel{\mathbf{R}_{F}} Q'$.
%\end{itemize}
We let $ \sim_F $ denote the largest  bio-simulation and we say that $P$ is \emph{bio-similar} to $Q$, with respect to $F$, if $P \sim_F Q$.
\end{definition}

\begin{remark}
An alternative way to look at a bio-simulation that respects $F$ is to define it as an ordinary bisimulation over the transition system labelled over $\{F,\neg F\}$ obtained by transforming each transition $P \xrightarrow{v} P'$ such that $v\entails F$ into $P \xrightarrow{F} P'$ and  each transition $P \xrightarrow{v} P'$ such that $v\not\entails F$ into $P \xrightarrow{\neg F} P'$.
\end{remark}

@@

Finally, we extend the classical result establishing the correspondence between the logical equivalence induced by HML with bisimilarity for proving that bio-similarity coincides with bio-logical equivalence.
